---
title: "Import gene sets from GMT files"
author:
  - name: Axel Klenk
    affiliation: Dept. of Medicine and Life Sciences, Universitat Pompeu Fabra, Barcelona
    email: axelvolker.klenk@upf.edu
  - name: Robert Castelo
    affiliation: Dept. of Medicine and Life Sciences, Universitat Pompeu Fabra, Barcelona
    email: robert.castelo@upf.edu
date: "`r Sys.Date()`" 
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    fig_caption: yes
    toc: true
pkgdown:
  as_is: true
bibliography: GSVAEuroBioC2024.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Import gene sets from GMT files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Vignette built using `GSVA`, version `r packageDescription("GSVA")[["Version"]]`.

# What are GMT files

An important source of gene sets is the
[Molecular Signatures Database (MSigDB)](https://www.gsea-msigdb.org/gsea/msigdb)
[@subramanian2005gene], which stores them in plain text files following the
so-called
[_gene matrix transposed_ (GMT) format](https://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29).
In the GMT format, each line stores a gene set with the following values
separated by tabs:

  * A unique gene set identifier.
  * A gene set description.
  * One or more gene identifiers.

Because each different gene set may consist of a different number of genes, each
line in a GMT file may contain a different number of tab-separated values. This
means that the GMT format is not a tabular format, and therefore cannot be directly
read with base R functions such as `read.table()` or `read.csv()`.

# How to read GMT files

We need a specialized function to read GMT files. We can find such a function in the
[GSEABase](https://bioconductor.org/packages/GSEABase) package with `getGmt()`, or
in the [qusage](https://bioconductor.org/packages/qusage) package with `read.gmt()`.

GSVA also provides such a function called `readGMT()`, which takes as first argument
the filename or URL of a, possibly compressed, GMT file. The call below illustrates
how to read a GMT file from MSigDB providing its URL. Note that we also load the
package 
[GSEABase](https://bioconductor.org/packages/GSEABase) because, by default, the value
returned by `readGMT()` is a `GeneSetCollection` object defined in that package.

```{r, message=FALSE, warning=FALSE}
library(GSEABase)
library(GSVA)

URL <- "https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2023.2.Hs/c7.immunesigdb.v2023.2.Hs.symbols.gmt"
genesets <- readGMT(URL, geneIdType=SymbolIdentifier())
class(genesets)
length(genesets)
genesets
```
If we have problems accessing that URL, we can also load a compressed version of this
GMT file, stored in the workshop package, as follows.

```{r}
fname <- system.file("extdata", "c7.immunesigdb.v2023.2.Hs.symbols.gmt.gz",
                     package="GSVAEuroBioC2024")
genesets <- readGMT(fname, geneIdType=SymbolIdentifier())
class(genesets)
length(genesets)
genesets
```
Next to the filename of the GMT file, we provided the argument
`geneIdType=SymbolIdentifier()`. This argument adds the necessary metadata that
later on allows other software to figure out what kind of gene identifiers are
used in this collection of gene sets, to attempt mapping them to other type of
identifiers, if necessary. While this argument is optional, we should always
try to provide it.

# Dealing with duplicated gene set names

The specification of the [GMT format](https://software.broadinstitute.org/cancer/software/gsea/wiki/index.php/Data_formats#GMT:_Gene_Matrix_Transposed_file_format_.28.2A.gmt.29)
establishes that duplicated gene set names are not allowed. For this reason,
the `getGmt()` function from the
[GSEABase](https://bioconductor.org/packages/GSEABase) package prompts an error
when duplicated gene names are found, while the `read.gmt()` function from the
[qusage](https://bioconductor.org/packages/qusage) package silently accepts them
in a list with duplicated element names.

The GSVA `readGMT()` function deals with duplicated gene set names as follows.
By default, `readGMT()` warns the user about a duplicated gene set name and
keeps only the first occurrence of the duplicated gene set in the returned
object. We can illustrate this situation with a GMT file from the MSigDB
database that happens to have duplicated gene set names and which we have
stored as part of this workshop package.

```{r, message=FALSE, error=TRUE}
fname <- system.file("extdata", "c2.all.v7.5.symbols.gmt.gz",
                     package="GSVAEuroBioC2024")
genesets <- getGmt(fname, geneIdType=SymbolIdentifier())
```
We can see that `getGmt()` prompts an error. We can see below that this does
not happen with `readGMT()` and that, by default, all but the first occurrence of the
duplicated gene set have been removed.

```{r}
genesets <- readGMT(fname, geneIdType=SymbolIdentifier())
genesets
any(duplicated(names(genesets)))
```
The parameter `deduplUse` in the `readGMT()` function allow one to apply other
policies to deal with duplicated gene set names, see its help page with
`?readGMT`.

# Returning a `list` instead of a `GeneSetCollection`

Storing gene sets in a `GeneSetCollection` object offers a number of advantages,
including the addition of metadata that enables an automatic mapping of the gene
identifiers that form the gene sets. For this reason, `readGMT()` returns such
an object by default. However, if a user prefers using gene sets stored as a
`list` object of `character` vectors, this is possible by setting the argument
`valueType="list"` in the call to `readGMT()`.

```{r}
genesets <- readGMT(fname, valueType="list")
class(genesets)
any(duplicated(names(genesets)))
head(lapply(genesets, head))
```

# Session information

```{r}
sessionInfo()
```

# References
